\chapter{Enumeration of programs}
\newcommand{\pr}{\mathcal{PR}}
The objective is to define an \emph{``effective'' enumeration} of URM
programs and URM-computable functions. These results will be
fundamental for our theory, and in particular for
\begin{itemize}
\item prove the existence of non computable functions (this proof will
  be similar to the first one we've seen in
  \S\ref{se:existence-non-2}, but with this method we'll be able to
  build them)
\item $S^m_n$ theorem (\textit{smn} theorem, or paramether theorem)
  \item universal function/machine.
\end{itemize}

Review: Let $ A, B $ be two sets,
\begin{itemize}
\item $ |A| = |B| $ if $ \exists f:A\rightarrow B $ biunivocal.

\item $ |A| \leq |B| $ if $ \exists f:A\rightarrow B $ injective or
  there exists $g$ opposite direction surjective.

\item if $|A| \leq |B|$ and $|B| \leq |A| \Rightarrow |A|=|B|$.
  Assuming the axiom of choice, if we have $\{A_i\}_{i \in I}$ family
  of non-empty sets $\forall i \in I \; A_i \neq \emptyset$, then
  there exists a function $$f:I \rightarrow \cup_{i \in I}A_i \quad
  \text{ s.t. }  \forall i \in I f(i) \in A_i$$
\end{itemize}

$A$ is countable if $ |A| \leq |\nat| $, that is,
$ \exists g: \nat \rightarrow A $ surjective. in this case we say that
$f$ is an enumeration of $X$, intuitively because we can enumerate all
elements in $X$ with \[f(0), f(1), f(2), \dots \]

An enumeration is \emph{without repetitions} if in addition to being
surjective it is also injective.

Let's say that an enumeration $f: \nat \rightarrow X$ is
\emph{effective} when both $f$ and $f^\prime$ are effective. When
$X = \nat^k \quad f^{-1}: X \rightarrow \nat$ can be proved to be
computable, otherwise we will use the intuitive notion of effective
function.

% when it is computable or
% made from pieces that are computable (e.g. ennuple of results).

\begin{lemma}
  The following sets are effectively enuerable:
  \begin{enumerate}[label=(\arabic*)]
  \item $ \nat^2 $
  \item $ \nat^3 $
  \item $\bigcup_{k\geq 1} \nat^k $
  \end{enumerate}
  \begin{proof}
    \begin{enumerate}[label=(\arabic*)]
    \item we alredy saw that \[\Pi : \nat^2 \rightarrow
        \nat\] \[\Pi(x,y) = 2^x(2y+1)-1\] Both are computable
      ($\in \pr$) with inverse
      \[\Pi^{-1} : \nat \rightarrow \nat^2\]
      \[\Pi^{-1}(x) = (\Pi_1(x), \Pi_2(x))\]
      \[\Pi_1, \Pi_2 \in \pr \Rightarrow \text{ computable}\]
    \item{
        We can consider \[\nu : \nat^3 \rightarrow
          \nat \] \[\nu(x,y,z)= \Pi(\Pi(x,y),z) \quad \nu \in \pr\] with
        the inverse built upon projections
        \[\nu^{-1} : \nat \rightarrow \nat^3\]
        \(\nu^{-1}(x) = (\nu_1(x), \nu_2(x), \nu_3(x))\)

        \(\nu_1(x) = \Pi_1(\Pi_1(x)), \nu_2(x) = \Pi_2(\Pi_2(x)),
        \nu_3(x) = \Pi_3(\Pi_3(x))\)

        Since $\nu_1, \nu_2, \nu_3 \in \pr$ are computable $\nu^{-1}$
        is computable }
    \item{ Observe that the encoding
        \[(a_1, \dots, a_k) \mapsto \prod^k_{i=1}p_i^{a_i+1}\] can't
        be used since it is injective, but not surjective (the
        codomain does not contain $0,1$ and any number $x$
        s.t. \(\exists i,j \; i<j \quad p_i \times x \wedge p_j | x\))

        The idea it to leverage the uniqueness of the binary
        rapresentation of natural numbers (positional).
        \[\tau : \cup_{k \geq 1} \nat^k \rightarrow \nat \]
        \[\tau (a_1, \dots, a_k ) = 2^{a_1}+2^{a_1+a_2+1}+ \dots + 2^{a_1+\dots + a_k + (k-1)} - 1\]
        \[\tau (a_1, \dots, a_k ) = \sum^k_{i=1}2^{(\sum_{j=1}^i a_j)
            + (i-1)}-1\]
        \begin{itemize}
        \item $\tau$ is intuitively effective (it uses only
          exclusively functions), but the domain doesn't allow us to
          prove its computability.
        \item $\tau$ is bijective: let $x \in \nat$, we need to find
          $(a_1, \dots, a_k)$ s.t. $\tau(a_1,\dots,a_k) = x$.
        \end{itemize}
      }
    \end{enumerate}
  \end{proof}
\end{lemma}

% everything after this was wrote by the professor, but is a lot less
% stuff for the same things in the italian notes, so I'll translate
% also the parts that he already translated here.
\textbf{Observation:} 

In particular the couples in $ \nat^2 $ can be encoded as
$ \pi(x,y) = 2^x(2y+1)-1 $ which is computable. The inverse is
$ \pi^{-1}(n) = (\pi_1(n), \pi_2(n)) $

The triple instead is a pair of a pair and an element.
$ \upsilon (x,y,z) = \pi (x, \pi(y,z))$. The inverse is also obtained
from the inverse of the first.

For lists we need an encoding
$ \tau . \bigcup_{K \geq 1} \nat^k \rightarrow \nat $ we exploit the
uniqueness of the prime numbers:
$ \tau(a_1,\dots,a_k) = \Pi_{i=1}^k p_i^{a_i}$ where $ p_i $ is the
$i$-th prime number. \\
This, however, leads us to lose any zeros, since
the encodings for (1,1) and (1,1,0) would be the same because the
exponential function in 0 = 1.

So we use something that works:
$ (\Pi_{i=1}^{k-1} p_i^{a_i}) \times p_k^{a_k+1} - 2$. For decoding we
can proceed as usual, but limited minimization can be used.\\ Hint:
$ max \{z \leq x . P(z)\} x - min\{\delta \leq x . P(x-\delta)\}$

But I don't just want the length, I also need the list of items. I
need a function:
\begin{equation*}
  a(x,i) = \begin{cases}
    (x+2)_i   & 1 \leq i \leq k-1 \\
    (x+2)_k-1 & i = k
  \end{cases}
\end{equation*}
And this is the inverse function of $\tau$. And these functions are
computable recursive primitives.

To compute instructions: Let us take $ \mathcal{F} $ set of URM
instructions, $ \mathcal{P} $ URM programs. Let's take function
$ \beta:\mathcal{F}\rightarrow\nat $
\begin{itemize}
\item $ \beta(z(n)) = 4 \times (n-1) $;
\item $ \beta(s(n)) = 4 \times (n-1)+1 $;
\item $ \beta(t(m,n)) = 4 \times \pi(m-1,n-1)+2 $;
\item $ \beta(j(m,n,t)) = 4 \times \upsilon(m-1,n-1,t-1)+3 $;
\end{itemize}

The decoding of this monstrosity is obtained from the previous inverse
functions applied on the basis of the dimension and the rest of the
number.

\textbf{Example:} P = T (1,2); S (2); T (2,1) = $ \tau(10,5,6) = 2^{10} 3^5 5^{6+1} -2 $

The two-way function between programs and numbers has been demonstrated.

\begin{notation} given an effective enumeration (in our case the one defined previously) we say that $ \gamma(P) $ is the code of P (also called G\"{o} of the number), if $ \gamma(P) = n $ then $P$ is the $n$-th program.
\end{notation}

\begin{notation} $ \Phi_n^{(k)}: \nat^k\rightarrow\nat $ function of $K$ arguments computed by program $n$, that is, by program $ \gamma^{-1}(n) $, if $k = 1$ is is omitted. The function domain  is $ W_n^{k} = dom(\Phi_k^{(k)}) = \{\vec{x} | \Phi_k^{(k)})(\vec{x})\downarrow  \} \subseteq \nat^k$

  The function codomain: $ E^{(k)}_n = \Phi_k^{(k)}) = \{\vec{x} | \vec{x} \in W_n^{(k)} \} \subseteq \nat^k$
\end{notation}

So for example the program $ \Phi_{19439999998} = x+1 $, $ W_{19439999998} = \nat $, $ E_{19439999998} = \nat \setminus \{0\} $

Now we have an enumeration of all the unary computable functions that is \{$ \Phi_n $ . $ n \in \nat $ \} each function with infinite repetitions.

Remember we have indicated the computable functions of $k$ arguments as $ \mathcal{C} ^ {(k)} $, where $ | \mathcal{C} ^ {(k)} | \leq (=) | \nat | $ and therefore being $ \mathcal{C} = \bigcup_{K \geq 1} \mathcal{C} ^ {(k)} $ a union of countable sets it is still countable.
