\chapter{Diagonalization}

Given a set $X$ we know that $ |X| \geq |2^X| $ is never valid, that is, the set of its parts is always bigger. Suppose there is $ f:X\rightarrow2^X $ surjective. Hence $ R = \{y . y \in X $ and $ x \not \in f(y) \ \} \in 2^X $, since f is surjective then $ \exists y_R \in X . f(y_R)  = R$. I consider the cases separately:
\begin{itemize}
\item $ y_R \in R $ then $f(y_R) \Rightarrow y_R \not \in R $
\item $ y_R \not \in R $ then $ f(y_R) \Rightarrow y_R \in R$
\end{itemize}
Now we prove that the set of the parts of the natural numbers is not countable. We assume there is a surjective function $ \nat \rightarrow 2^\nat $. This means that I can enumerate subsets $ X_i $ s.t. $ i \in \nat $. At this point I create a matrix where each row $i$ of column $j = 1$ if $ x_i \in X_j $ and 0 otherwise. Now consider the inverted diagonal, that is if $ x_i \not \in X_i $, in this way it is different from all the columns, that is $ \forall i . R \not= X_i $ because $ n \in R \Leftrightarrow b \not \in X_n $ absurd since I assumed that $ \{X_0 \dots X_n \} = 2^\nat$.

We conclude that $ |\nat| \not \geq |2^\nat| $

But we want $ |\nat| < |2^\nat| $.

But: $ |\nat| \leq |2^\nat| \land |\nat| \not\geq |2^\nat| \implies |\nat| < |2^\nat| $

I take the set of the characteristic functions $g$ of $ 2^\nat $ and I call it $Y$, I take the set of all the functions $ \mathcal{F} $, of course it holds that $ Y \subseteq \mathcal{F} $ and therefore $ |Y| \leq |\mathcal{F}| $ but being that $ |\nat < |2^\nat| $ then I also have that $ |\nat| \leq |\mathcal{F}| $

There is a total function that cannot be computed. We know how to enumerate computable functions because we can enumerate them in the form of numbers, repeating some of them. Now let's define $ f(n) $ as non-total computable.

Matrix where the columns are the functions computed by the program $ i \in \nat $, that is $ \{\phi_i . i \in \nat \} $ and the rows are the arguments 0,1, \dots.

\begin{equation*}
  f(n) = \begin{cases}
    \phi_n(n)+1 & $ se $\phi(n)\downarrow \\
    0           & $ se $\phi(n)\uparrow
  \end{cases}
\end{equation*}

we observe that $f$ is total by construction and $ f \not= \phi_n \forall n, n \in \nat $ Furthermore by construction it is different from all computable functions and therefore it is not computable.
